---
title: "Hormuz Shock Tracker"
subtitle: "Reading the 2026 Iran–US war through the DHL Global Connectedness lens"
author: "Bhavesh Gupta"
date: last-modified
date-format: "D MMMM YYYY, HH:mm 'UTC'"
format:
  html:
    code-fold: true
    fig-width: 9
    fig-height: 5
abstract: |
  On 28 February 2026 the United States and Israel struck Iranian nuclear and
  missile infrastructure. Within 48 hours the Strait of Hormuz — carrier of
  roughly one fifth of global crude and a large share of LNG — was closed to
  commercial traffic. Brent broke $100 per barrel within ten days. This site
  tracks the ongoing economic shock the way the DHL Initiative on Globalization
  tracks connectedness: empirically, reproducibly, and in a form a policymaker
  can read in five minutes. Every figure is generated from an open data pipeline
  (R + DuckDB + Quarto); the warehouse and the CI workflow that refreshes it
  are in the [GitHub repository](https://github.com/pmxlr8/DS_R_Geopolitics).
---

```{r setup}
#| include: false
suppressPackageStartupMessages({
  library(DBI); library(duckdb); library(dplyr); library(tidyr)
  library(ggplot2); library(scales); library(plotly); library(readr)
  library(lubridate); library(crosstalk); library(htmltools)
})
source("R/db_helpers.R")
source("R/connectedness.R")

con <- db_connect(read_only = TRUE)

events <- db_read(con, "SELECT * FROM events ORDER BY date")
prices <- db_read(con, "SELECT * FROM v_price_shock ORDER BY date") |>
  mutate(date = as.Date(date))
depth  <- db_read(con, "SELECT * FROM v_trade_depth ORDER BY iso3c, year")

WAR_START <- as.Date("2026-02-28")

theme_shock <- function() {
  theme_minimal(base_size = 13) +
    theme(plot.title       = element_text(face = "bold"),
          plot.subtitle    = element_text(color = "grey30"),
          panel.grid.minor = element_blank(),
          legend.position  = "top",
          legend.title     = element_blank())
}

# Plotly renders `geom_vline(xintercept = as.numeric(date))` with a
# time-zone offset on date axes — the line ends up half a day off. Using
# plotly's native `shapes` layout primitive avoids the conversion entirely.
war_shape <- list(
  type = "line", xref = "x", yref = "paper",
  x0 = WAR_START, x1 = WAR_START, y0 = 0, y1 = 1,
  line = list(color = "rgba(80,80,80,0.6)", dash = "dash", width = 1.2)
)
war_annot <- list(
  x = WAR_START, y = 1, xref = "x", yref = "paper",
  text = "28 Feb 2026 · strikes", showarrow = FALSE,
  xanchor = "left", yanchor = "top",
  font = list(size = 10, color = "grey40")
)
```

```{r header-stats}
#| echo: false
latest_date  <- max(prices$date, na.rm = TRUE)
latest_brent <- prices |> filter(series == "Brent", date == latest_date) |> pull(value)
latest_wti   <- prices |> filter(series == "WTI",   date == latest_date) |> pull(value)
latest_spread <- latest_brent - latest_wti
days_since_war <- as.integer(Sys.Date() - WAR_START)
render_ts <- format(Sys.time(), "%d %b %Y · %H:%M UTC", tz = "UTC")

stat_box <- function(label, value, sub = NULL) {
  div(style = "flex:1; padding:14px 18px; background:#f8f9fa; border-radius:8px; margin:4px;",
    div(style = "font-size:11px; text-transform:uppercase; letter-spacing:.05em; color:#6c757d;", label),
    div(style = "font-size:22px; font-weight:600; color:#212529; margin-top:4px;", value),
    if (!is.null(sub)) div(style = "font-size:11px; color:#6c757d; margin-top:2px;", sub)
  )
}

div(style = "display:flex; flex-wrap:wrap; margin:16px -4px 24px -4px;",
  stat_box("Brent, latest",
           sprintf("$%.2f / bbl", latest_brent),
           sub = format(latest_date, "%d %b %Y")),
  stat_box("Brent–WTI spread",
           sprintf("$%.2f", latest_spread),
           sub = "geopolitical risk premium"),
  stat_box("Days since strikes",
           sprintf("%d", days_since_war),
           sub = "from 28 Feb 2026"),
  stat_box("Site refreshed",
           render_ts,
           sub = "auto-rebuilt daily · 06:00 UTC")
)
```

::: {.callout-note appearance="simple"}
Data auto-refreshes daily via [GitHub Actions](https://github.com/pmxlr8/DS_R_Geopolitics/actions).
Every figure pulls live from World Bank WDI and FRED; source badges sit under each chart.
:::

## The shock, in one paragraph

Crude oil traded near $70 a barrel before the strikes. By 8 March 2026, Brent
broke $100 for the first time in four years; during the blockade peak it spiked
above $119. The IMF cut its 2026 MENA growth forecast from 3.9% to 1.1% and
warned that continued escalation could push the global economy into recession.
The US–Iran–Israel ceasefire signed on 7 April has been fraying: Iran cancelled
the reopening of the Strait on 17 April and the US blockade of Iranian ports
remains in force.

The rest of this page traces that shock through four measurements.

---

## Chart A — The spike

> Daily WTI and Brent spot prices. Use the legend to isolate a series, or drag
> on the chart to zoom into any window. The dashed line marks the 28 Feb 2026
> strikes.

```{r chart-a}
#| label: chart-a-spike
p <- ggplot(prices, aes(x = date, y = value, color = series,
                        text = paste0(format(date, "%d %b %Y"),
                                      "<br>", series, ": $",
                                      format(round(value, 2), nsmall = 2), "/bbl"))) +
  geom_line(aes(group = series), linewidth = 0.6) +
  scale_y_continuous(labels = dollar_format(suffix = "/bbl")) +
  scale_color_manual(values = c(WTI = "#2C7FB8", Brent = "#D95F02")) +
  labs(title = "Oil prices: the 2026 Iran-war spike",
       subtitle = "WTI and Brent daily spot (USD/bbl)",
       x = NULL, y = NULL) +
  theme_shock()

ggplotly(p, tooltip = "text") |>
  layout(shapes = list(war_shape), annotations = list(war_annot),
         legend = list(orientation = "h", y = 1.08),
         hovermode = "x unified")
```

*Source: FRED — [`DCOILWTICO`](https://fred.stlouisfed.org/series/DCOILWTICO), [`DCOILBRENTEU`](https://fred.stlouisfed.org/series/DCOILBRENTEU) · daily refresh*

The picture tells two stories. The first is speed: Brent added roughly forty
dollars in the first six trading sessions of March — faster than after Russia's
2022 invasion of Ukraine on the same axis. The second is the tail: prices
remained above the pre-war range into the April ceasefire, suggesting markets
were pricing not only the blockade but the probability it would resume.

---

## Chart B — The geopolitical risk premium

> The Brent–WTI spread is a classic read of Middle East risk: Brent prices
> Hormuz-adjacent crude; WTI prices US crude. When the spread widens, markets
> are charging extra for geopolitical exposure.

```{r chart-b}
#| label: chart-b-spread
spread <- prices |>
  pivot_wider(names_from = series, values_from = value) |>
  filter(!is.na(Brent), !is.na(WTI)) |>
  mutate(spread = Brent - WTI)

p <- ggplot(spread, aes(x = date, y = spread,
                        text = paste0(format(date, "%d %b %Y"),
                                      "<br>spread: $",
                                      format(round(spread, 2), nsmall = 2)))) +
  geom_hline(yintercept = 0, color = "grey70") +
  geom_line(aes(group = 1), color = "#7B3294", linewidth = 0.6) +
  scale_y_continuous(labels = dollar_format(suffix = "/bbl")) +
  labs(title = "The geopolitical risk premium",
       subtitle = "Brent minus WTI, daily",
       x = NULL, y = NULL) +
  theme_shock()

ggplotly(p, tooltip = "text") |>
  layout(shapes = list(war_shape), annotations = list(war_annot),
         hovermode = "x unified")
```

*Source: derived from FRED Brent and WTI daily spots.*

A normal Brent–WTI spread sits in the low single digits, reflecting quality and
logistics. During the 2026 shock it widened sharply as Hormuz-adjacent supply
became harder to replace. For policymakers the spread is a cleaner read than
the absolute price — it strips out demand-side noise and leaves the
geography-of-supply story exposed.

---

## Chart C — The rewiring (interactive)

> Trade depth (exports + imports as a share of GDP) for a panel of major
> economies. **Use the checkboxes below** to filter to the countries you care
> about.

```{r chart-c}
#| label: chart-c-rewiring
#| column: page
focus <- c("IRN", "SAU", "ARE", "TUR", "DEU", "CHN", "USA", "IND", "EUU")

panel <- depth |> filter(iso3c %in% focus)

sd_panel <- SharedData$new(panel, key = ~country, group = "ctry_filter")

p <- ggplot(sd_panel,
            aes(x = year, y = trade_depth_pct, color = country, group = country,
                text = paste0(country, " · ", year,
                              "<br>trade depth: ",
                              format(round(trade_depth_pct, 1), nsmall = 1), "%"))) +
  geom_line(linewidth = 0.7) +
  scale_y_continuous(labels = label_percent(scale = 1)) +
  scale_color_brewer(palette = "Set1") +
  labs(title = "Trade depth by economy, 2000–2024",
       subtitle = "Trade (goods + services) as % of GDP · click a country to isolate",
       x = NULL, y = NULL) +
  theme_shock() +
  theme(legend.position = "right")

bscols(
  widths = c(3, 9),
  filter_checkbox("ctry", "Filter countries", sd_panel, ~country, inline = FALSE),
  ggplotly(p, tooltip = "text") |> layout(hovermode = "closest")
)
```

*Source: World Bank WDI indicator [`NE.TRD.GNFS.ZS`](https://data.worldbank.org/indicator/NE.TRD.GNFS.ZS) · annual refresh.*

Three groups sit in different positions on this chart. Gulf economies (UAE,
Saudi Arabia) trade at 60–200% of GDP — their exposure is structural. Large
continental blocs (US, China, India) trade at 20–40% — exposure is real but
buffered by domestic demand. The European Union sits in between. A Hormuz
shock does not hit everyone the same way; the chart makes *semi-globalization*
— Ghemawat's observation that the world is neither fully flat nor fully local
— visible on one axis.

---

## Chart D — The connectedness score

> A z-score of each economy's latest trade depth against its own **2018–2022
> pre-war baseline**. Above zero = more connected than baseline; below zero =
> decoupling underway.

```{r chart-d}
#| label: chart-d-score
scored <- depth |>
  shock_score(baseline_start = 2018, baseline_end = 2022) |>
  filter(iso3c %in% focus) |>
  group_by(iso3c) |>
  filter(year == max(year[!is.na(shock_z)])) |>
  ungroup() |>
  filter(!is.na(shock_z)) |>
  arrange(shock_z) |>
  mutate(country = factor(country, levels = country))

p <- ggplot(scored,
            aes(x = country, y = shock_z, fill = shock_z > 0,
                text = paste0(country,
                              "<br>year scored: ", year,
                              "<br>z-score: ",
                              format(round(shock_z, 2), nsmall = 2),
                              "<br>baseline mean: ",
                              format(round(base_mean, 1), nsmall = 1), "%"))) +
  geom_col() +
  scale_fill_manual(values = c(`TRUE` = "#2C7FB8", `FALSE` = "#D95F02"),
                    guide = "none") +
  coord_flip() +
  labs(title = "Who is above or below their 2018–2022 connectedness baseline?",
       subtitle = "Standard deviations from 2018–2022 trade-depth mean, latest WDI year",
       x = NULL, y = "z-score") +
  theme_shock()

ggplotly(p, tooltip = "text")
```

*Source: derived from WDI indicator [`NE.TRD.GNFS.ZS`](https://data.worldbank.org/indicator/NE.TRD.GNFS.ZS); methodology on [this page](pages/methodology.qmd).*

The bar chart compresses the four-chart story into one line-per-country read.
Economies pushed below zero are already registering a structural pullback from
trade relative to their 2018–2022 average. Those above zero are either late to
reflect the shock in annual data or benefiting from reconfiguration (e.g.,
transshipment winners). Because WDI data is annual, the post-February-2026
effects will not appear here in full until the 2026 release lands — another
reason the oil-market charts above carry the near-real-time signal.

---

## Timeline

Event annotations throughout this page are loaded from a version-controlled
YAML file (`data/events.yml`) so the war itself is diff-able in git.

```{r timeline}
#| echo: false
events |>
  mutate(date = format(as.Date(date), "%Y-%m-%d")) |>
  select(date, category, label, source) |>
  knitr::kable(col.names = c("Date", "Category", "Event", "Source"))
```

```{r cleanup}
#| include: false
db_disconnect(con)
```

---

*See also: [historical analogs](pages/historical_analogs.qmd) ·
[methodology](pages/methodology.qmd) ·
[data sources](pages/data_sources.qmd).*
