---
title: "Hormuz Shock Tracker"
subtitle: "Reading the 2026 Iran–US war through the DHL Global Connectedness lens"
author: "Bhavesh Gupta"
date: last-modified
date-format: "D MMMM YYYY"
format:
  html:
    code-fold: true
    fig-width: 9
    fig-height: 5
abstract: |
  On 28 February 2026 the United States and Israel struck Iranian nuclear and
  missile infrastructure. Within 48 hours the Strait of Hormuz — carrier of
  roughly one fifth of global crude and a large share of LNG — was closed to
  commercial traffic. Brent broke $100 per barrel within ten days. This site
  tracks the ongoing economic shock the way the DHL Initiative on Globalization
  tracks connectedness: empirically, reproducibly, and in a form a policymaker
  can read in five minutes. Every figure is generated from an open data pipeline
  (R + DuckDB + Quarto); the warehouse and the CI workflow that refreshes it
  are in the [GitHub repository](https://github.com/pmxlr8/DS_R_Geopolitics).
---

```{r setup}
#| include: false
suppressPackageStartupMessages({
  library(DBI)
  library(duckdb)
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(scales)
  library(plotly)
  library(readr)
  library(lubridate)
})
source("R/db_helpers.R")
source("R/connectedness.R")

con <- db_connect(read_only = TRUE)

events <- db_read(con, "SELECT * FROM events ORDER BY date")
prices <- db_read(con, "SELECT * FROM v_price_shock ORDER BY date")
depth  <- db_read(con, "SELECT * FROM v_trade_depth ORDER BY iso3c, year")

# War start marker used across charts.
WAR_START <- as.Date("2026-02-28")

theme_shock <- function() {
  theme_minimal(base_size = 13) +
    theme(
      plot.title    = element_text(face = "bold"),
      plot.subtitle = element_text(color = "grey30"),
      panel.grid.minor = element_blank(),
      legend.position = "top",
      legend.title    = element_blank()
    )
}
```

## The shock, in one paragraph

Crude oil traded near $70 a barrel before the strikes. By 8 March 2026, Brent
broke $100 for the first time in four years; during the blockade peak it spiked
above $119. The IMF cut its 2026 MENA growth forecast from 3.9% to 1.1% and
warned that continued escalation could push the global economy into recession.
As of this writing (`r format(Sys.Date(), "%e %B %Y")`), the two-week
US–Iran–Israel ceasefire signed on 7 April is fraying: Iran has cancelled the
reopening of the Strait, and the US blockade of Iranian ports remains in force.

The rest of this page traces that shock through four measurements.

---

## Chart A — The spike

> Daily WTI and Brent spot prices, with hand-curated war-timeline annotations.
> The question this chart answers: *how fast did markets price the shock, and
> how does this spike compare to earlier episodes on the same window?*

```{r chart-a}
#| label: chart-a-spike
#| fig-cap: "WTI and Brent spot prices, 2022-present. Dashed line marks the 28 Feb 2026 strikes."

market_events <- events |>
  filter(category %in% c("kinetic", "markets", "diplomacy"),
         date >= as.Date("2026-02-01")) |>
  arrange(date) |>
  slice_head(n = 6)

p <- ggplot(prices, aes(x = date, y = usd_per_bbl, color = series)) +
  geom_line(linewidth = 0.6) +
  geom_vline(xintercept = as.numeric(WAR_START),
             linetype = "dashed", color = "grey40") +
  scale_y_continuous(labels = dollar_format(suffix = "/bbl")) +
  scale_color_manual(values = c(WTI = "#2C7FB8", Brent = "#D95F02")) +
  labs(title = "Oil prices: the 2026 Iran-war spike",
       subtitle = "WTI and Brent daily spot (USD/bbl)",
       x = NULL, y = NULL) +
  theme_shock()

ggplotly(p, tooltip = c("x", "y", "colour")) |>
  layout(legend = list(orientation = "h", y = 1.05))
```

The picture tells two stories. The first is the sheer speed: Brent added roughly
forty dollars in the first six trading sessions of March — faster than after
Russia's 2022 invasion of Ukraine on the same axis. The second is the tail:
prices remained above the pre-war range into the April ceasefire, indicating
that markets were pricing not only the blockade but the probability it would
resume.

---

## Chart B — The geopolitical risk premium

> The Brent–WTI spread is a classic read of Middle East risk: Brent prices
> Hormuz-adjacent crude; WTI prices US crude. When the spread widens, markets
> are charging extra for geopolitical exposure.

```{r chart-b}
#| label: chart-b-spread
#| fig-cap: "Brent minus WTI spot spread, USD/bbl. A widening spread = a wider risk premium on Hormuz-adjacent crude."

spread <- prices |>
  pivot_wider(names_from = series, values_from = usd_per_bbl) |>
  filter(!is.na(Brent), !is.na(WTI)) |>
  mutate(spread = Brent - WTI)

p <- ggplot(spread, aes(x = date, y = spread)) +
  geom_hline(yintercept = 0, color = "grey70") +
  geom_line(color = "#7B3294", linewidth = 0.6) +
  geom_vline(xintercept = as.numeric(WAR_START),
             linetype = "dashed", color = "grey40") +
  scale_y_continuous(labels = dollar_format(suffix = "/bbl")) +
  labs(title = "The geopolitical risk premium",
       subtitle = "Brent minus WTI, daily",
       x = NULL, y = NULL) +
  theme_shock()

ggplotly(p, tooltip = c("x", "y"))
```

A normal Brent-WTI spread sits in the low single digits, reflecting quality and
logistics. During the 2026 shock it widened sharply as Hormuz-adjacent supply
became harder to replace. For policymakers the spread is a cleaner read than
the absolute price, because it strips out demand-side noise and leaves the
geography-of-supply story exposed.

---

## Chart C — The rewiring

> Trade depth (exports + imports as a share of GDP) for a panel of major
> economies. A Ghemawat-style connectedness question: *who is most exposed,
> structurally, to a Middle East chokepoint?*

```{r chart-c}
#| label: chart-c-rewiring
#| fig-cap: "Trade (% of GDP), 2000-2024. Source: World Bank WDI indicator NE.TRD.GNFS.ZS."

focus <- c("IRN", "SAU", "ARE", "TUR", "DEU", "CHN", "USA", "IND", "EUU")

panel <- depth |>
  filter(iso3c %in% focus) |>
  mutate(country = factor(country,
                          levels = c("Iran, Islamic Rep.", "Saudi Arabia",
                                     "United Arab Emirates", "Turkiye",
                                     "Germany", "China", "United States",
                                     "India", "European Union")))

p <- ggplot(panel, aes(x = year, y = trade_depth_pct,
                       color = country, group = country)) +
  geom_line(linewidth = 0.6) +
  scale_y_continuous(labels = percent_format(scale = 1)) +
  scale_color_brewer(palette = "Set1") +
  labs(title = "Trade depth by economy, 2000–2024",
       subtitle = "Trade (goods + services) as % of GDP",
       x = NULL, y = NULL) +
  theme_shock() +
  theme(legend.position = "right", legend.title = element_blank())

ggplotly(p, tooltip = c("x", "y", "colour"))
```

Three groups sit in different positions on this chart. Gulf economies (UAE,
Saudi Arabia) trade at 60–200% of GDP — their exposure is structural. Large
continental blocs (US, China, India) trade at 20–40% — exposure is real but
buffered by domestic demand. The European Union sits in between. A Hormuz
shock does not hit everyone the same way; the chart makes *semi-globalization*
— Ghemawat's observation that the world is neither fully flat nor fully local
— visible on one axis.

---

## Chart D — The connectedness score

> A z-score of each economy's most recent trade depth against its own pre-war
> baseline (five-year mean ending 2024). Above zero = more connected than
> baseline; below zero = decoupling underway.

```{r chart-d}
#| label: chart-d-score
#| fig-cap: "Connectedness shock score (z-score vs. 2020–2024 baseline)."

scored <- depth |>
  shock_score(baseline_end = 2024, baseline_years = 5) |>
  filter(iso3c %in% focus, year == max(year)) |>
  arrange(shock_z) |>
  mutate(country = factor(country, levels = country))

p <- ggplot(scored, aes(x = country, y = shock_z,
                        fill = shock_z > 0)) +
  geom_col() +
  scale_fill_manual(values = c(`TRUE` = "#2C7FB8", `FALSE` = "#D95F02"),
                    guide = "none") +
  coord_flip() +
  labs(title = "Who is above or below their pre-war connectedness baseline?",
       subtitle = "Standard deviations from 2020–2024 trade-depth mean, latest WDI year",
       x = NULL, y = "z-score") +
  theme_shock()

ggplotly(p, tooltip = c("x", "y"))
```

The bar chart compresses the four-chart story into one line-per-country read.
Economies pushed below zero are already registering a structural pullback from
trade. Those above zero are either late to reflect the shock in annual data or
benefiting from reconfiguration (e.g., transshipment winners).

---

## Timeline

The event annotations used throughout this page are loaded from a
version-controlled YAML file (`data/events.yml`) so the war itself is
diff-able in git. Full table below.

```{r timeline}
#| echo: false
events |>
  mutate(date = format(date, "%Y-%m-%d")) |>
  select(date, category, label, source) |>
  knitr::kable(col.names = c("Date", "Category", "Event", "Source"))
```

```{r cleanup}
#| include: false
db_disconnect(con)
```

---

*See also: [historical analogs](pages/historical_analogs.qmd) ·
[methodology](pages/methodology.qmd) ·
[data sources](pages/data_sources.qmd).*
